Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-10 manuscript/supplemental_material.Rnw
fix species table and add more expression figures

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("magrittr")
library("tidyr")

@@

\begin{table}
  \begin{tabular}{c p{1.5in} c p{2in}}
    \toprule
    Species                       & Collector & Experiment Number & Sample Number(s) \\
    \midrule
    \textit{Ateles fusciceps}     & Kurt Benirschke & & \\
    \textit{Bos taurus}           & Bob Kreft & & \\
    \textit{Canis familiaris}     & Dalen Agnew & & (Bounty Hubler) \\
    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& & \\
    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
    \textit{Loxodonta africana}   & Kurt Benirschke & & \\
    \textit{Homo sapiens}         & Nandor Gabor, \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
                                             & SRR1850945,
                                               SRR1850946, 
                                               SRR1850951, 
                                               SRR1850952, 
                                               SRR1850953, 
                                               SRR1850954, 
                                               SRR1850957, 
                                               SRR1850958, 
                                               SRR1850965, 
                                               SRR1850966, 
                                               SRR1850975, 
                                               SRR1850976, 
                                               SRR1850983, 
                                               SRR1850984, 
                                               SRR1850987, 
                                               SRR1850988, 
                                               SRR1850991, 
                                               SRR1850992, 
                                               SRR1850993, 
                                               SRR1850994, 
                                               SRR1850995, 
                                               SRR1850996, 
                                               SRR1851003, 
                                               SRR1851004, 
                                               SRR1851011, 
                                               SRR1851012 \\
    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & & \\
    \textit{Mus musculus}         & Juan Gonzalez, \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
                                     & SRR243594,
                                       SRR243595,
                                       SRR243596,
                                       SRR243597,
                                       SRR243598,
                                       SRR243599,
                                       SRR243600,
                                       SRR243601,
                                       SRR243602,
                                       SRR243586,
                                       SRR243587,
                                       SRR243588,
                                       SRR243589,
                                       SRR243590,
                                       SRR243591,
                                       SRR243592,
                                       SRR243593\\
    \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
    \textit{Pan paniscus}         & William Mintz Fields & & (Teco)\\
    \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\
    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016467 \\
    \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\
    \bottomrule
  \end{tabular}
  \caption{Species, experiment and species numbers used in the analyses presented in this paper}
  \label{tab:srr_numbers}
\end{table}

@@

\begin{figure}
<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
star.logs.unmapped <- 
    star.logs[grepl("%",field) & 
                  grepl("unmapped",field) & 
                      !grepl("too many mismatches",field),]
star.logs.unmapped <- 
    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
print(ggplot(star.logs.unmapped,
             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
      + geom_boxplot()
      + facet_grid(field~.,scales="free")
      + ylab("Percent of Reads")
      + xlab(NULL)
      + theme_bw()
      + theme(legend.position="top",
              plot.margin=unit(c(0,0,0,1.1),"strwidth",data=list("","","","e8")),
              axis.title.x=element_blank(),
              axis.text.x = element_blank()),#element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
      vp=viewport(layout.pos.col=1,layout.pos.row=1)
      )
pushViewport(viewport(layout.pos.col=1,layout.pos.row=1))
grid.text("A",
          y=unit(1,"npc")-unit(0.1,"cm"),
          x=unit(0,"npc")+unit(0.1,"cm"),
          gp=gpar(fontsize=24),
          just=c("left","top"))
popViewport()          
print(ggplot(star.logs[grepl("input reads$",field),],
             aes(x=capfirst(species),y=as.numeric(value)))
      + geom_boxplot()
      + scale_y_log10()
      + ylab("Reads")
      + xlab("Species")
      + theme_bw()
      + theme(legend.position="top",
              plot.margin=unit(c(0,1.6,0,0),"strheight",data=list("","TO","","")),
              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
      vp=viewport(layout.pos.col=1,layout.pos.row=2))
pushViewport(viewport(layout.pos.col=1,layout.pos.row=2))
grid.text("B",
          y=unit(1,"npc")-unit(0.1,"cm"),
          x=unit(0,"npc")+unit(0.1,"cm"),
          gp=gpar(fontsize=24),
          just=c("left","top"))
popViewport(2)
@ 
\caption{RNAseq read and alignment statistics for all species. A)
  Boxplot of percentages of unmapped reads in each species; 0\% of
  reads in all species were unmapped because of mismatches. B) Boxplot
  of number of RNAseq reads in each species.}
\label{fig:transcriptome_statistics}
\end{figure}

@@

@@

>>>>

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.65\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("magrittr")
library("tidyr")

@@

\begin{table}
  \raggedright
  \begin{tabular}{c >{\raggedright}p{1.5in} c >{\raggedright\arraybackslash}p{2in}}
    \toprule
    Species                       & Collector & Experiment & Sample(s) \\
    \midrule
    \textit{Ateles fusciceps}     & Kurt Benirschke & & \\
    \textit{Bos taurus}           & Bob Kreft & & \\
    \textit{Canis familiaris}     & Dalen Agnew & & (Bounty Hubler) \\
    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& & \\
    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
    \textit{Loxodonta africana}   & Kurt Benirschke & & \\
    \textit{Homo sapiens}         & \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
                                             & SRR1850945,
                                               SRR1850946, 
                                               SRR1850951, 
                                               SRR1850952, 
                                               SRR1850953, 
                                               SRR1850954, 
                                               SRR1850957, 
                                               SRR1850958, 
                                               SRR1850965, 
                                               SRR1850966, 
                                               SRR1850975, 
                                               SRR1850976, 
                                               SRR1850983, 
                                               SRR1850984, 
                                               SRR1850987, 
                                               SRR1850988, 
                                               SRR1850991, 
                                               SRR1850992, 
                                               SRR1850993, 
                                               SRR1850994, 
                                               SRR1850995, 
                                               SRR1850996, 
                                               SRR1851003, 
                                               SRR1851004, 
                                               SRR1851011, 
                                               SRR1851012 \\
    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & & \\
    \textit{Mus musculus}         & \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
                                     & SRR243594, 
                                       SRR243595, 
                                       SRR243596, 
                                       SRR243597, 
                                       SRR243598, 
                                       SRR243599, 
                                       SRR243600, 
                                       SRR243601, 
                                       SRR243602, 
                                       SRR243586, 
                                       SRR243587, 
                                       SRR243588, 
                                       SRR243589, 
                                       SRR243590, 
                                       SRR243591, 
                                       SRR243592, 
                                       SRR243593\\
    \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
    \textit{Pan paniscus}         & William Mintz Fields & & (Teco)\\
    \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\
    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016466 \\
    \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\
    \bottomrule
  \end{tabular}
  \caption{Species, experiment and species numbers used in the analyses presented in this paper}
  \label{tab:srr_numbers}
\end{table}

@@

\begin{figure}[hb]
<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
star.logs.unmapped <- 
    star.logs[grepl("%",field) & 
                  grepl("unmapped",field) & 
                      !grepl("too many mismatches",field),]
star.logs.unmapped <- 
    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
print(ggplot(star.logs.unmapped,
             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
      + geom_boxplot()
      + facet_grid(field~.,scales="free")
      + ylab("Percent of Reads")
      + xlab(NULL)
      + theme_bw()
      + theme(legend.position="top",
              plot.margin=unit(c(0,0,0,1.1),"strwidth",data=list("","","","e8")),
              axis.title.x=element_blank(),
              axis.text.x = element_blank()),#element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
      vp=viewport(layout.pos.col=1,layout.pos.row=1)
      )
pushViewport(viewport(layout.pos.col=1,layout.pos.row=1))
grid.text("A",
          y=unit(1,"npc")-unit(0.1,"cm"),
          x=unit(0,"npc")+unit(0.1,"cm"),
          gp=gpar(fontsize=24),
          just=c("left","top"))
popViewport()          
print(ggplot(star.logs[grepl("input reads$",field),],
             aes(x=capfirst(species),y=as.numeric(value)))
      + geom_boxplot()
      + scale_y_log10()
      + ylab("Reads")
      + xlab("Species")
      + theme_bw()
      + theme(legend.position="top",
              plot.margin=unit(c(0,1.6,0,0),"strheight",data=list("","TO","","")),
              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
      vp=viewport(layout.pos.col=1,layout.pos.row=2))
pushViewport(viewport(layout.pos.col=1,layout.pos.row=2))
grid.text("B",
          y=unit(1,"npc")-unit(0.1,"cm"),
          x=unit(0,"npc")+unit(0.1,"cm"),
          gp=gpar(fontsize=24),
          just=c("left","top"))
popViewport(2)
@ 
\caption{RNAseq read and alignment statistics for all species. A)
  Boxplot of percentages of unmapped reads in each species; 0\% of
  reads in all species were unmapped because of mismatches. B) Boxplot
  of number of RNAseq reads in each species.}
\label{fig:transcriptome_statistics}
\end{figure}

@@

\begin{figure}[hb]
<<adam12.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="ADAM12" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=5,
                 axis.text.size=5,
                 fontsize=8
                 )
@ 
\caption{\genetreeexplanation{\textit{ADAM12}}}
\label{fig:adam12_evolution_tree}
\end{figure}

\begin{figure}
<<kiss1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="KISS1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=5,
                 axis.text.size=5,
                 axis.text.angle=0,
                 fontsize=5
                 )
@ 
\caption{\genetreeexplanation{\textit{KISS1}}}
\label{fig:kiss1_evolution_tree}
\end{figure}

\begin{figure}
<<gpc3.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="GPC3" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=5,
                 axis.text.size=5,
                 fontsize=8
                 )
@ 
\caption{\genetreeexplanation{\textit{GPC3}}}
\label{fig:gpc3_evolution_tree}
\end{figure}

@@

\begin{table}[hb]
  \footnotesize
<<human.lineage.specific,echo=FALSE,results="asis">>=
hum.all.lin.spec <- placenta.lineage.specific.table("hum.all")
setkey(hum.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed in \human
  as compared to all other analyzed species.}
\label{tab:human_lineage_specific}
\end{table}
